BOLD Parameter Estimation Using Particle Filters

### 2010-08-16 doc/thesis/conclusion.tex
small edits

One significant finding of this work, that would not be clear without 
calculating a full posterior distribution, is the interplay 
between the parameters. The results of simulations clearly demonstrate
that identifying a single set of parameters is not possible with this 
model. Although sensitivity tests in Deneux et al. certainly hinted
at this, the current work clearly demonstrates this fact \cite{Deneux2006}. Therefore,
any single estimate of parameters is insufficient for analysis. As such,
identification of the BOLD model that do not treat the parameters as distributions
will not be able to overcome the inherent indeterminability of the
parameters. Besides the Unscented Kalman Filter, this is the only approach
that accomplishes this task without repeatedly calculating the parameters 
to build a distribution. The Unscented Kalman Filter, of course, is limited
to Gaussian estimation which limits its ability to estimate the posterior.

The primary reason this method has not been used before is the high 
dimensionality of the system. In Murray, 2008 the idea of learning
the parameters is floated as the correct method, yet 
learning 7 parameters with a Monte Carlo method was deemed intractable
\cite{Murray2008}. The 
concern was that so many parameters creates a prohibitively large 
search space, that requires too many particle to learn properly. To overcome
this difficulty, instead of starting the algorithm with 1000 particles,
the initial number of particles was set to 16,000. This meant that
when the search space was the largest, the number of particles was
sufficiently dense to represent the prior distribution. Then resampling
was performed for the first time, the number of particles was dropped, 
since the weight of most the particles had dropped to 0. The result is
a particle filter algorithm that spends computing resources only where
they are really needed.

Many of the methods to determining areas of neural activation aim
to be Bayesian, yet ultimately must make limiting assumptions about
the distributions that remove many of the benefits. 
The particle filter is a true Bayesian non-parametric algorithm that, given
enough measurements will approach the true probability distribution
of the parameters. While the conclusion that the parameters are not
uniquely identifiable is a disappointing one in light years of research 
attempting to learn these them, it in fact further demonstrates
the need to estimate probability distribution rather than a single 
parameter estimate. At the same time, the output of the particle filter
is still able to give good estimates of the BOLD output, and is not
dependent on heavy Gaussian smoothing. Therefore, all the current 
experimental paradigms to determine regional activity are still viable.
To borrow a computer term, the particle filter algorithm is backward 
compatible with the current methods. Indeed, the particle filter excelled
at determining activation in very low SNR simulations 
(\autoref{sec:Multi-voxel Simulation}). Not only that, but the particle
filter algorithm identified areas of activation that were completely
missed by SPM (\autoref{fig:comp6}).

>>>>

One significant finding of this work, that would not be clear without 
calculating a full posterior distribution, is the interplay 
between the parameters. The results of simulations clearly demonstrate
that identifying a single set of parameters is not possible with this 
model. Although sensitivity tests in Deneux et al. certainly hinted
at this, the current work clearly demonstrates this fact \cite{Deneux2006}. Therefore,
any single estimate of parameters is insufficient for analysis. As such,
identification of the BOLD model that do not treat the parameters as distributions
will not be able to overcome the inherent ill-posed nature of the
parameters. Besides the Unscented Kalman Filter, this is the only approach
that accomplishes this task without repeatedly calculating the parameters 
to build a distribution. The Unscented Kalman Filter though, is restricted
to Gaussian estimation which limits its ability to estimate the posterior.

The primary reason this method has not been used before is the high 
dimensionality of the system. In Murray, 2008 the idea of learning
the parameters is floated as the correct method, yet 
learning 7 parameters with a Monte Carlo method was deemed intractable
\cite{Murray2008}. The 
concern was that so many parameters creates a prohibitively large 
search space, which requires too many particle to learn properly. To overcome
this difficulty, instead of starting the algorithm with 1000 particles,
the initial number of particles was set to 16,000. This meant that
when the search space was the largest, the number of particles was
sufficiently dense to represent the prior distribution. The result is
a particle filter algorithm that spends computing resources only where
it is really needed.

The particle filter is a true Bayesian non-parametric algorithm that, given
enough measurements will approach the true probability distribution
of the parameters. The conclusion that the parameters are not
uniquely identifiable is a disappointing one in light of the plethora
of research papers that have attempted to learn them. However, the fact
that the parameters are ill-posed further demonstrates
the need to estimate probability distribution rather than a single 
parameter estimate. At the same time, the output of the particle filter
is still able to give good estimates of the BOLD output, and is not
dependent on heavy Gaussian smoothing. Therefore, all the current 
experimental paradigms to determine regional activity are still viable.
To borrow a computer term, the particle filter algorithm is backward 
compatible with the current methods. Indeed, the particle filter excelled
at determining activation in very low SNR simulations 
(\autoref{sec:Multi-voxel Simulation}). Not only that, but the particle
filter algorithm identified areas of activation that were completely
missed by SPM (\autoref{fig:comp6}).

The physiological plausibility is also of great value to future
research. Unlike traditional methods, use of the BOLD model allows 
the incorporation of outside, physiological knowledge in regression.
If, for instance, it becomes possible to measure $v$ in vivo that data
could instantly be added to the model to improve regression. Data can
also be retrieved from the model. BOLD parameter distributions could 
for the first time be used as diagnosis tools. In the past the fact 
that the parameters are
ill-posed would have prevented such research. However, now that a full
distribution is available, comparing parameter estimates is actually
feasible.

### 2010-08-16 doc/thesis/discussion.tex
small edits

\chapter{Discussion}
\label{sec:Discussion}
\section{Review of Results}
% Overview of results
%% Parameters under-constrained
%% Output estimates good
There are two possible causes for the large variance in
parameter estimates. First, it could be that the particle filter is
incapable of learning the model. However, given the high quality
of BOLD estimates, this is almost certainly not the case. The other 
possible explanation is that the given stimulus is not sufficiently
varied to bring out all the properties of the system. While it is 
possible that this is the case, it is unlikely that even the best 
stimulus sequence could differentiate all the parameters. The fact
that \autoref{sec:VeryLongSim}, which used randomly spaced impulses,
still had a high correlation definitely indicates that the parameters
are ill-defined.

@@

\chapter{Future Work}
\label{sec:FutureWork}
% Enhancements/Dehancements 
%% prior
%%% flatten prior
%%% Larger variance
%% Detrending
%%% \% difference using spline?
%%% "smart knots"
%%% DC gain as a parameter
%%% linearizing
%% Experimental Changes
%%% Variations in stimuli
%%% longer timeseries or artificial
%%% Automatic estimation of measurement error.
\section{Algorithm Improvements}
\label{sec:Particle Filter Variations}
There are a few areas where future research may improve upon the current
work. The first is modifying the prior 
distribution. The prior distribution used in this work is listed in \autoref{tab:Prior}, 
and is based on the findings of Friston et al. \cite{Friston2000}. Unfortunately, that
result depended on a quadratic approximation (Volterra Series) which has
not been extensively tested (at least not in a published work). 
Additionally, the current prior is
based only around the BOLD output, which is why it is good at generating
estimates of the BOLD signal, but weak in estimating parameters. 
The best solution would be to have
in-vivo estimates of the actual parameters, although this
is unlikely to happen. Instead, better estimates could be found
using population studies with additional measurements as discussed in 
\autoref{sec:Sideways Measurements}. Better knowledge of the prior distribution
would make it possible to decrease variance of certain 
parameters, reducing the mobility of the parameters.

Another major area for improvement is removing drift. 
A large number of detrending methods were tested, but every method was in some way limited. 
One viable method that could be utilized, given the right experimental design, is detrending
based on areas of low activity. This has the advantage that it wouldn't require an arbitrary
constant to be added to the pre-processed signal for the BOLD model to fit properly.
The disadvantage of this approach is that it could hide long fall times by normalizing them
out. It also requires periodic breaks in the stimulus, which could reduce value of those
samples for fitting purposes.

In terms of dealing with drift, linearizing is another possibility.
This would use the delta between measurements for fitting rather than the 
direct value. This has the advantage of not requiring detrending and thus 
gives nearly raw data to the particle filter
for processing. The effectiveness of this method depends directly on the type of stimulus.
In a test with rapid impulse stimuli, this could be extremely effective because
the DC level has minimal data; conversely prolonged flat level levels make this less effective.
In tests I found that large drift-low white noise
type signals performed much better with a linearization approach, as one might expect. 
For the stimulus sequence used in \autoref{sec:RealData}, the results tended to be worse
than using spline detrending.

@@

\subsection{Experimental Changes}
\label{sec:Sideways Measurements}
One definite way of improving the results of the particle filter is additional
measurements. While increasing the sample rate of FMRI scanners may
not be possible, simultaneous measurements of volume and flow is
possible, albeit at 9T in a cat \cite{Hu2009}. Just one of those measurements
though would be extremely powerful when incorporated into the BOLD
model. By adding another measurement, the
variance in the parameter estimates would significantly drop. The fact
that such a measurement would be closer to the true stimulus would make it
all the more powerful. 
A more conventional method of adding measurements is to 
simply perform longer FMRI tests. Although this wouldn't be 
groundbreaking, it would certainly increase the ability of the 
particle filter to develop the posterior distribution. 
It may also be possible to concatenate multiple FMRI tests together.
It should be possible to link the multiple runs simply through
the final parameter distribution from the previous run.

>>>>

\chapter{Discussion}
\label{sec:Discussion}
\section{Review of Results}
% Overview of results
%% Parameters under-constrained
%% Output estimates good
There are two possible causes for the large variance in
parameter estimates. First, it could be that the particle filter is
incapable of learning the model. However, given the high quality
of BOLD estimates, this is almost certainly not the case. The other 
possible explanation is that the given stimulus is not sufficiently
varied to bring out all the properties of the system. While it is 
possible that this is the case, it is unlikely that any
stimulus sequence could differentiate all the parameters. 
Its also extremely intesting that the histograms of parameters' mean
approach the width of the prior, even when the particle filter converges
proprerly to far thinner distributions. Its possible the prior 
distribution is far wider than 
it needs to be. Finally, the fact that \autoref{sec:VeryLongSim}, which 
used randomly spaced impulses,
still had a high correlation further indicates that the parameters
are ill-defined.

@@

\chapter{Future Work}
\label{sec:FutureWork}
% Enhancements/Dehancements 
%% prior
%%% flatten prior
%%% Larger variance
%% Detrending
%%% \% difference using spline?
%%% "smart knots"
%%% DC gain as a parameter
%%% linearizing
%% Experimental Changes
%%% Variations in stimuli
%%% longer timeseries or artificial
%%% Automatic estimation of measurement error.
\section{Algorithm Improvements}
\label{sec:Particle Filter Variations}
There are a few areas where future research may improve upon the current
work. The first is modifying the prior 
distribution. The prior distribution used in this work is listed in \autoref{tab:Prior}, 
and is based on the findings of Friston et al. \cite{Friston2000}. Unfortunately, that
result depended on a quadratic approximation (Volterra Series) which has
not been extensively tested (at least not in a published work). 
This is in additoin to the fact that the parameters are ill-defined, and
thus the parameters found in Friston et al. are most likely far too wide.
The prior is also
based completely on the BOLD output, which is why it is good at generating
estimates of the BOLD signal, but weak at estimating parameters. 
The best solution would be to have
in-vivo estimates of the actual parameters, although this
is unlikely to happen. Instead, better estimates could be found
using population studies with additional measurements as discussed in 
\autoref{sec:Sideways Measurements}. Better knowledge of the prior distribution
would make it possible to decrease variance of certain 
parameters, reducing the mobility of the parameters.

Another major area for improvement is removing drift. 
A large number of detrending methods were tested, but every method was in some way limited. 
One potentially viable method that could be utilized, given the right 
experimental design, is detrending based on intervals of low activity. This 
has the advantage that it wouldn't require an arbitrary
constant to be added to the pre-processed signal for the
BOLD model to fit properly.  The disadvantage of this approach is 
that it could hide long fall times by normalizing them
out. It also requires periodic breaks in the stimulus, 
which could reduce value of those samples for fitting purposes.

In terms of dealing with drift, linearizing is another possibility.
This would use the delta between measurements for fitting, rather than the 
direct value. This has the advantage of not requiring detrending and thus 
gives nearly raw data to the particle filter for processing. 
The effectiveness of this method depends directly on the type of stimulus.
In a test with rapid impulse stimuli, this could be extremely effective because
the DC level has minimal data; conversely prolonged flat levels make this less effective.
In tests I found that large drift/low white noise
type signals performed much better with a linearization approach.
For the stimulus sequence used in \autoref{sec:RealData}, the results tended to be worse
than using spline detrending.

@@

\section{Experimental Design Improvements}
\label{sec:Sideways Measurements}
One definite way of improving the results of the particle filter is through
additional measurements. While increasing the sample rate of FMRI scanners may
not be possible, simultaneous measurements of volume and flow is
possible, albeit at 9T in a cat \cite{Hu2009}. Just one of those measurements
though would be extremely powerful when incorporated into the BOLD
model. By adding another measurement, the
variance in the parameter estimates would significantly drop. The fact
that such a measurement would be closer to the true stimulus would make it
all the more powerful. 
A more conventional method of adding measurements is to 
simply perform longer FMRI tests. Although this wouldn't be 
groundbreaking, it would certainly increase the ability of the 
particle filter to develop the posterior distribution. 
It may also be possible to concatenate multiple FMRI tests together.
It should be possible to link the multiple runs simply through
the final parameter distribution from the previous run.

### 2010-08-16 doc/thesis/results.tex
small edits

\begin{figure}
\centering
\subfigure[Region labels for simulated slice.]{
\includegraphics[height=9.5cm]{images/simregions.png} \label{fig:simslice_mask} }
\subfigure[Grey matter map for simulate region.]{\includegraphics[height=9.5cm]{images/sim_gm}}
\subfigure[Residual Map. Scale is Normalized error. Lower (yellow) is better.]
{\label{fig:simslice_hm_res} \includegraphics[height=9cm]{images/sim_hm}}
\subfigure[Mutual Information Map. Scale is bits. Higher (yellow) is better.]
{\label{fig:sim_hm_mi} \includegraphics[height=9cm]{images/sim_hm_mi}}
\caption{Comparison of activation with greymatter, parameter regions.}
\label{fig:simslice_hm}
\end{figure}

@@

The histograms again demonstrate that a single point estimate of the parameters 
is elusive for this set of parameters. However the data clearly show the power
of the particle filter at identifying regions of activation.
 The thresholds applied to this
slice, both for mutual information and residuals are arbitrary. Applying a
threshold is helpful for visualization, but is rarely useful for further analysis.
As noted in \autoref{sec:SingleVoxelReview},
the false positives present in the Mutual Information
map are different from those in the residual map. This furthers the argument 
for combining the two metrics to increase power. Although
at first glance it would appear that there are false negatives in 
\autoref{fig:sim_hm_mi}; this is
not actually the case. POSSUM simulates different tissues, and white matter
does not typically  have a BOLD response. This explains some of the holes in regions
2 and 3. These results certainly indicate that the particle filter is effective
at regressing against a noisy signal.

>>>>

\begin{figure}
\centering
\subfigure[Region labels for simulated slice.]{
\includegraphics[height=9cm]{images/simregions.png} \label{fig:simslice_mask} }
\subfigure[Grey matter map for simulate region.]{\includegraphics[height=9cm]{images/sim_gm}}
\subfigure[Residual Map. Scale is Normalized error. Lower (yellow) is better.]
{\label{fig:simslice_hm_res} \includegraphics[height=8cm]{images/sim_hm}}
\subfigure[Mutual Information Map. Scale is bits. Higher (yellow) is better.]
{\label{fig:sim_hm_mi} \includegraphics[height=8cm]{images/sim_hm_mi}}
\caption{Comparison of activation with greymatter, parameter regions.}
\label{fig:simslice_hm}
\end{figure}

@@

The histograms again demonstrate that a single point estimate of the parameters 
is elusive for this set of parameters. However the data clearly show the power
of the particle filter at identifying regions of activation.
The thresholds applied to this slice, both for mutual information and 
residuals somewhat arbitrary. Applying a
threshold is helpful for visualization but is rarely useful for further analysis.
As noted in \autoref{sec:SingleVoxelReview},
the false positives present in the Mutual Information
map are different from those in the residual map. This furthers the argument 
for combining the two metrics to increase power. Although
at first glance it would appear that there are false negatives in 
\autoref{fig:sim_hm_mi}; this is
not actually the case. POSSUM simulates different tissues, and white matter
does not typically  have a BOLD response. This explains some of the holes in regions
2 and 3. These results certainly indicate that the particle filter is effective
at regressing against a noisy signal.
